Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-03-21 paper/paper.tex
touchup axioms, fix crefs, determinization disambiguation

% -*- fill-column: 80; -*-
\documentclass[review,anonymous,screen,acmsmall,nonacm]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{xspace}

@@

\newcommand{\Subst}{\textrm{Subst}}
\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\semg}[1]{\sem{#1}\gamma}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\textbf{String}}
\newcommand{\Char}{\textbf{Char}}
\newcommand{\stringg}{\texttt{String}}
\newcommand{\charg}{\mathtt{Char}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\semcat}{\mathbf{C}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathtt{Bool}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

\newtheorem{axiom}{Axiom}[section]
\newtheorem{construction}{Construction}

@@

Our Agda prototype is based on a \emph{denotational semantics} of
\theoryabbv. The core idea of the denotational semantics is
\cite{elliottSymbolicAutomaticDifferentiation2021}, Elliott describes
a formal grammar as type-level predicates on strings that prove
language membership. That is, a \emph{formal grammar} $A$ is a
function $\String \to \Set$ such that for a string $w$, $A~w$ is the
set of ``proofs'' showing that $w$ belongs to the language recognized
by $A$. We show that all linear types in \theoryabbv can be so
interpreted as an abstract formal grammar in this sense, and that
linear terms are a kind of \emph{parse transformer}, a function that
takes a parse tree from one grammar to a parse tree in a different
grammar but over the same underlying string.

@@

In addition to these type-theoretic principles, we need two additional
axioms that do not generally hold in systems based on linear
logic.

\paragraph{Distributivity}
First, we need that additive conjunction \emph{distributes}
over additive disjunction --- e.g., in the finitary case that $0 \& A
\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most general form,

@@

Our primary use of distributivity is to define an equivalence that
expresses a linear type $A$ as a sum over which character it starts with, if any,
\(
A \cong \left(A \& I \right) \oplus \LinSigTyLimit{c}{\Sigma_0}{A \& \left( \literal{c} \otimes \top \right)}
\). We use this equivalence when building a parser for the traces of the lookahead
automaton in \cref{fig:binop-inductive}.

\begin{corollary}
  \label{cor:binary-mono}
  Distributivity implies that the coprojections $\inl : A \to A \oplus B$,
  $\inr : B \to A \oplus B$ of a binary sum are monomorphisms.
\end{corollary}

\paragraph{Disjointness of Constructors}
Second, we need that the different constructors of
$\bigoplus$ are disjoint. That is, we want to enforce that $\sigma~x~a$ and
$\sigma~x'~a'$ of type $\LinSigTyLimit{x}{X}{A\,x}$ are not equal whenever $x \neq x'$.
However, because these are linear terms we cannot state their disequality
directly. Instead, we encode the disequality via a function out of an equalizer,

\begin{axiom}[$\sigma$-Disjointness]
\label{ax:disjointness}
For any $A : X \to L$ and $x \neq x' : X$ there is a function,
\[
  \uparrow(\equalizer{b}{\sigma\,x \circ \pi_1}{\sigma\,x'\circ \pi_2}
  \lto 0)
\]
  where $b: A(x) \& A(x')$. That is, the grammar of pairs of an
$a:A(x)$ and an $a': A(x')$ such that $\sigma\,x\,a = \sigma\,x'\,a'$
is empty.
\end{axiom}

\steven{need to get cref to cite this as a lemma}
We use $\sigma$-disjointness in \cref{lem:unambig-to-disjoint} to proof that
unambiguous sums have disjoint summands.

@@

So far, our calculus is a somewhat generic combination of dependent
types with non-commutative linear types. In order to carry out formal
grammar theory and define parsers, we need only add a few
grammar-specific constructions. First for each character in our
alphabet, we add a corresponding linear type $c$. We can then define a
non-linear type $\Char$ as the disjunction of all of these characters,
and define a type $\String$ as the Kleene star of $\Char$, i.e. as an
inductive linear type.
Then we add a function $\mathsf{read} :
\ltonl{(\top\lto \String)}$ that intuitively ``reads'' the input
string from the input and makes it available. It is important that the
input type of $\mathsf{read}$ is $\top$, which can control any amount
of resources, and not $\I$ which controls no resources. Further, we
add an axiom that $\lambda s. \mathsf{read}(!(s)) \equiv \lambda
s. s$ where $!$ is the unique function $\ltonl(\String\lto \top)$,
i.e., that if we have a string, but then throw it away and read
it from the input, then in fact that is equivalent to the string we
were given originally. This ensures that the elements of the $\String$
type always stand for the actual input string in our reasoning. In the
next section, we will show how these basic principles are enough to
provide a basis for verified parsing and formal grammar theory.

@@

%% \paragraph{Weak and Strong Equivalence}
In the theory of formal grammars, there are two different notions of
equivalence: up to weak generative capacity, meaning just which strings are
accepted by the grammar; and up to \emph{strong} generative capacity, when
the parse trees of the two grammars are isomorphic
\cite{chom1963}. Using linear types as grammars, we can define both of
these notions of equivalence in \theoryabbv.
\begin{definition}
  \label{def:weakequiv}
  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exist
  parse transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B
    \lto \A)}$. $\A$ is a \emph{retract} of $\B$ if $\A$ and $\B$ are
  weakly equivalent and $\lambda a. g(f(a)) \equiv \lambda a.a$. They
  are \emph{strongly equivalent} if further the other composition is
  the identity, i.e., $\lambda b. f(g(b)) \equiv \lambda b.b$.
\end{definition}

%% That is, grammars are weakly equivalent if we can construct a parse
%% tree of $B$ from a parse tree of $A$ and vice-versa, whereas they are
%% strongly equivalent if these processes are inverse to each other. The
%% definition of retract is a useful intermediate notion. In retrospect,
%% our axiom about $\String$ says that $\String$ is a retract of $\top$.

@@

%% \paragraph{Ambiguity}
A formal grammar $\A$ is ambiguous if there are multiple parse trees
for the \emph{same} string $\w$.  For example, $\a \oplus \a$ is
ambiguous because there are two parses of $\stringquote{a}$,
constructed using $\inl$ and $\inr$. On the other hand, a formal
grammar is unambiguous when there is at most one parse tree for any
input string. We can capture this notion as a type in \theoryabbv in a
clever way:
\begin{definition}
  \label{def:unambig}
  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
  then $\f \equiv \g$.
\end{definition}
\cref{def:unambig} can be read more intuitively as stating that $\A$
is unambiguous if there is at most one way to transform parses of any
other grammar $\B$ into parses of $\A$. This notion of an unambiguous
type is the analog for linear types of the definition of a (homotopy)
\emph{proposition} in the terminology of homotopy type
theory\cite{hottbook}. The most basic unambiguous types are $\top$ and
$0$, and in a system of classical logic all unambiguous types would
have to be equivalent to one of these, but with our axioms we can show
also that $\I$ and literals $\literal c$ are unambiguous. To see this, first, we
establish two useful properties of unambiguity.
\begin{lemma}[\agdalogo]
  If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
  As a consequence of \steven{fix cref} \cref{cor:binary-mono}, if a binary disjunction $A \oplus B$ is unambiguous then
  $A$ and $B$ are each unambiguous.
\end{lemma}
From the first principle, we can prove that $\String$ is unambiguous,
since it is a retract of $\top$. In fact, observe that if $\A$ is a
retract of $\B$ and $\B$ is unambiguous, then in fact $\A$ and $\B$
are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
\lambda b. b$ follows because $\B$ is unambiguous. Therefore
$\String$ is also strongly equivalent to $\top$. Next, since $\String$ is
defined as a Kleene star, we can easily show that $\String \cong \I
\oplus \Char \oplus (\Char \otimes \Char \otimes \String)$. Then by
the second principle, we have that $\I$ and
$\Char$ and each literal
$\literal c$ are unambiguous as well.

% \paragraph{A Grammar of Strings}
% Define the character grammar $\charg$ to be the disjunction over characters in the alphabet $\Sigma$,
% $\charg = \LinSigTy {c} {\Sigma} c$. Then define a linear type of strings as a
% linear list of characters,
% $\stringg = \charg^{*}$.

% In the grammars model, we can build a
% strong equivalence between $\top$ and $\stringg$. hen building parsers, we may wish to enforce this
% syntactically by including $\top \cong \stringg$ as an axiom.

% Through inclusion of this axiom, $! : \stringg \to \top$ is an isomorphism, and
% thus also a monomorphism. That is, $\stringg$ is unambiguous.
% \max{we should have a section of the previous section which gives the axioms, this one and the distributivity axiom}

% \paragraph{Parseability}
% We say $A$ is \emph{parseable} if there exists a grammar $B$ and a map
% $\top \to A \oplus B$. Taking $\top \cong \stringg$, a grammar $A$ is parseable
% exactly when we may read an arbitrary string into a parse of $A$ or a $B$-valued
% refutation.

@@

We now turn to our main task, which is using our linear type system to
implement verified parsers. Given a grammar defined as a linear type
$A$, a first attempt at defining a parser would be to implement a
function $\uparrow{(\String \lto A)}$. But since our linear functions
must be total, this means that we can construct an $A$ parse for
\emph{every} input string, which is impossible for most grammars of
interest. Instead we might try to write a partial function as a
$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option''
monad. This allows for the possibility that the input string doesn't
parse, but is far too weak as a specification: we can trivially
implement a parser for any type by always returning $\inr$. The
correct notion of a parser should be one that allows for failure, but
only in the case that a parse cannot be constructed.

\begin{definition}
  Linear types $A$ and $B$ are \emph{disjoint} if there is a function
  $\ltonl {\left( A \& B \lto 0 \right)}$.
\end{definition}

\begin{definition}
  A parser for a linear type $A$ is the choice a type $A_{\neg}$ disjoint from
  $A$ and function $\uparrow{(\String \lto
    A \oplus A_{\neg})}$.
\end{definition}
Here we replace $\top$ in our partial parser type with a type
$A_{\neg}$ that we can think of as a negation of $A$. The function
$\uparrow{(A \& A_{\neg} \lto 0)}$ ensures that it is impossible for
$A$ and $A_{\neg}$ to parse the same input string. This means that in
defining a parser, we will need to define a kind of negative grammar
for strings that do not parse.

@@

\steven{
\begin{lemma}[\agdalogo]
  \label{lem:unambig-to-disjoint}
  If $\LinSigTyLimit{x}{X}{A\,x}$ is unambiguous, then for $x \neq x'$, $A\,x$ and $A,\y$ are
  disjoint.

  In particular, if the binary product $A \oplus A_{\neg}$ is unambiguous, then
  $A$ and $A_{\neg}$ are disjoint.
\end{lemma}
}

Writing a parser as a linear term in this way is an intrinsic
verification of the \emph{soundness} of the parser completely for free
from the typing: any $\inl$ parse that we return \emph{must}
correspond to a parse tree of the input string. Further if we verify
the disjointness property we then also get the \emph{completeness} of
the parser as well, that it never fails to generate an $A$ parse when
it is possible.

Our main method for constructing verified parsers is to show that a
grammar $A$ is weakly equivalent to a grammar for a deterministic
automaton. Parsers for deterministic automata are simple to implement
by stepping through the states of the automaton, with the rejecting
traces serving as the negative grammar. This is sufficient due to the
following:
\begin{lemma}[\agdalogo]
  \label{lem:wk-eqv-parse}
  If $\A$ is weakly equivalent to $\B$ then any parser for $\A$ can be
  extended to a parser for $\B$.
\end{lemma}
Here we need both directions of the weak equivalence. We need $A \lto
B$ to extend the parser from $\String \lto A \oplus A_{\neg}$ to
$\String \lto B \oplus A_{\neg}$ but then we also need $B \lto A$ to
establish that $A_{\neg}$ is disjoint from $B$.

@@

\newcommand{\states}{\mathtt{states}}
\newcommand{\labelt}{label}
\newcommand{\transitions}{\texttt{transitions}}
\newcommand{\epstransitions}{\epsilon\texttt{transitions}}
\newcommand{\isAcc}{\mathtt{isAcc}}
\newcommand{\init}{\mathtt{init}}
\newcommand{\src}{\mathtt{src}}
\newcommand{\dst}{\mathtt{dst}}
\newcommand{\parse}{\mathtt{parse}}
\newcommand{\print}{\mathtt{print}}
\newcommand{\epssrc}{\epsilon\mathtt{src}}
\newcommand{\epsdst}{\epsilon\mathtt{dst}}

\newcommand{\N}{\texttt{N}}

@@

Because DFAs are deterministic, we are able to prove that their type
of traces are unambiguous and define a parser for them. In particular
we show that for any start state $s$, ${\LinSigTy {b} {\Bool}
  \Trace_{D}~s~b}$ is a retract of $\stringg$. That is, first we
construct a function $\parse_{D}$ that is a parser for
$\Trace_D~{s}~{true}$, with $\Trace_D~{s}~{false}$ being the disjoint
type used. The fact that these types are disjoint follows by showing
that this function is part of a retraction, i.e., that there is only
one way to trace through a deterministic automaton.

The parser,
$\parse_{D}$,
is defined by recursion on strings in \cref{fig:printdfa}. If this string is empty, then $\parse_{D}$
defines a linear function that terminates a trace at the input state $s$. If the
string is nonempty, then $\parse_{D}$ walks forward in $D$ from the input state $s$ by
the character at the head of the string.
%
The inverse, $\print_{D}$ is defined by recursion on traces. If the trace is defined
via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
is defined by $\cons$ then $\parse_{D}$ appends the character from the most
recent transition to the output
string and recurses.
%
We prove this is a retraction by induction on traces.

\begin{theorem}[\agdalogo]
  $\parse_D s$ is a parser for $\Trace_D~s~\true$.
\end{theorem}

@@

To construct a term from DFA traces to NFA traces, we similarly strengthen our
induction hypothesis and build a parse transformer
\[
  \DtoN : \ltonl {\left( \Trace_{\D}~\X~\true \lto \bigoplus\limits_{s : \N.\mathsf{states}}\bigoplus\limits_{\texttt{sInX} : \X \ni \s} \Trace_{\N}~s~\true \right)}
\]

\steven{To define $\DtoN$, we take the smallest trace through $\N$ subject to an
  ordering on its states and transitions, where the ordering is supplied
  non-linearly. In essence, the given ordering specifies a disambiguation strategy.}

@@

\subsection{Unrestricted Grammars}
\newcommand{\stringNL}{String'}

>>>>

% -*- fill-column: 80; -*-
\documentclass[review,anonymous,screen,acmsmall,nonacm]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{xspace}
\usepackage{thmtools} % needed for sane cleveref naming of theorems

@@

\newcommand{\Subst}{\textrm{Subst}}
\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\semg}[1]{\sem{#1}\gamma}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\StringSem}{\textbf{String}}
\newcommand{\CharGram}{\texttt{Char}}
\newcommand{\StringGram}{\texttt{String}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\semcat}{\mathbf{C}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathtt{Bool}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

\newcommand{\states}{\mathtt{states}}
\newcommand{\labelt}{label}
\newcommand{\transitions}{\texttt{transitions}}
\newcommand{\epstransitions}{\epsilon\texttt{transitions}}
\newcommand{\isAcc}{\mathtt{isAcc}}
\newcommand{\init}{\mathtt{init}}
\newcommand{\src}{\mathtt{src}}
\newcommand{\dst}{\mathtt{dst}}
\newcommand{\pparse}{\mathtt{parse}}
\newcommand{\print}{\mathtt{print}}
\newcommand{\epssrc}{\epsilon\mathtt{src}}
\newcommand{\epsdst}{\epsilon\mathtt{dst}}
\newcommand{\balanced}{\mathtt{balanced}}
\newcommand{\Dyck}{\mathtt{Dyck}}
\newcommand{\N}{\texttt{N}}

% These declarations are to get cleveref to play nice with
% different theorem-like environemnts. Otherwise \cref{lem1.1}
% will display 'Theorem 1.1'

\declaretheoremstyle[headfont=\normalfont\itshape]{defstyle}
\declaretheoremstyle[headfont=\normalfont\scshape]{thmstyle}

\declaretheorem[name=Definition, style=defstyle, numberwithin=section]{definition}
\declaretheorem[style=thmstyle, name=Theorem, numberlike=definition]{theorem}
\declaretheorem[style=thmstyle, name=Axiom, numberlike=definition]{axiom}
\declaretheorem[style=thmstyle, name=Corollary, numberlike=definition]{corollary}
\declaretheorem[style=thmstyle, name=Construction, numberlike=definition]{construction}

\crefname{definition}{Definition}{Definitions}
\crefname{theorem}{Theorem}{Theorems}
\crefname{axiom}{Axiom}{Axioms}
\crefname{corollary}{Corollary}{Corollarys}
\crefname{construction}{Construction}{Constructions}

% \newtheorem{axiom}{Axiom}[section]
% \newtheorem{construction}{Construction}

@@

Our Agda prototype is based on a \emph{denotational semantics} of
\theoryabbv. The core idea of the denotational semantics is
\cite{elliottSymbolicAutomaticDifferentiation2021}, Elliott describes
a formal grammar as type-level predicates on strings that prove
language membership. That is, a \emph{formal grammar} $A$ is a
function $\StringSem \to \Set$ such that for a string $w$, $A~w$ is the
set of ``proofs'' showing that $w$ belongs to the language recognized
by $A$. We show that all linear types in \theoryabbv can be so
interpreted as an abstract formal grammar in this sense, and that
linear terms are a kind of \emph{parse transformer}, a function that
takes a parse tree from one grammar to a parse tree in a different
grammar but over the same underlying string.

@@

In addition to these type-theoretic principles, we need two additional
axioms that do not generally hold in systems based on linear
logic. First, we need that additive conjunction \emph{distributes}
over additive disjunction --- e.g., in the finitary case that $0 \& A
\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most general form,

@@

\begin{corollary}
  \label{cor:binary-mono}
  Distributivity implies that the coprojections $\inl : A \to A \oplus B$,
  $\inr : B \to A \oplus B$ of a binary sum are monomorphisms --- i.e. if
  $\inl\,a \equiv \inl\,a'$, then $a \equiv a'$.
\end{corollary}

\steven{cite reference for distributivity implies coprojections mono. Cockett
  ``Introduction to Distributive Categories''? Awodey?}

Our primary use of distributivity is to define an equivalence that
expresses a linear type $A$ as a sum over which character it starts with, if any,
\(
A \cong \left(A \& I \right) \oplus \LinSigTyLimit{c}{\Sigma_0}{\left( A \& \left( \literal{c} \otimes \top \right) \right)}
\). We use this equivalence when building a parser for the traces of the lookahead
automaton in \cref{fig:binop-inductive}.

Second, we need that the different constructors of
$\bigoplus$ are disjoint. That is, we want to enforce that $\sigma~x~a$ and
$\sigma~x'~a'$ of type $\LinSigTyLimit{x}{X}{A\,x}$ are not equal whenever $x \neq x'$.
However, because these are linear terms we cannot state their disequality
directly. Instead, we encode the disequality via a function out of an equalizer,

\begin{axiom}[$\sigma$-Disjointness]
\label{ax:disjointness}
For any $A : X \to L$ and $x \neq x' : X$ there is a function,
\[
  \uparrow(\equalizer{b}{\left( \sigma\,x \circ \pi_1 \right)}{\left(  \sigma\,x'\circ \pi_2 \right)}
  \lto 0)
\]
  where $b: A(x) \& A(x')$. That is, the grammar of pairs of an
$a:A(x)$ and an $a': A(x')$ such that $\sigma\,x\,a = \sigma\,x'\,a'$
is empty.
\end{axiom}

We use $\sigma$-disjointness in \cref{lem:unambig-to-disjoint} to prove that
unambiguous sums have disjoint summands.

@@

So far, our calculus is a somewhat generic combination of dependent
types with non-commutative linear types. In order to carry out formal
grammar theory and define parsers, we need only add a few
grammar-specific constructions. First for each character in our
alphabet, we add a corresponding linear type $c$. We can then define a
non-linear type $\CharGram$ as the disjunction of all of these characters,
and define a type $\StringGram$ as the Kleene star of $\CharGram$, i.e. as an
inductive linear type.
Then we add a function $\mathsf{read} :
\ltonl{(\top\lto \StringGram)}$ that intuitively ``reads'' the input
string from the input and makes it available. It is important that the
input type of $\mathsf{read}$ is $\top$, which can control any amount
of resources, and not $\I$ which controls no resources. Further, we
add an axiom that $\lambda s. \mathsf{read}(!(s)) \equiv \lambda
s. s$ where $!$ is the unique function $\ltonl(\StringGram\lto \top)$,
i.e., that if we have a string, but then throw it away and read
it from the input, then in fact that is equivalent to the string we
were given originally. This ensures that the elements of the $\StringGram$
type always stand for the actual input string in our reasoning. In the
next section, we will show how these basic principles are enough to
provide a basis for verified parsing and formal grammar theory.

@@

%% \paragraph{Weak and Strong Equivalence}
In the theory of formal grammars, there are two different notions of
equivalence: up to weak generative capacity, meaning just which strings are
accepted by the grammar; and up to \emph{strong} generative capacity, when
the parse trees of the two grammars are isomorphic
\cite{chom1963}. Using linear types as grammars, we can define both of
these notions of equivalence in \theoryabbv.

\begin{definition}
  \label{def:weakequiv}
  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exist
  parse transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B
    \lto \A)}$. $\A$ is a \emph{retract} of $\B$ if $\A$ and $\B$ are
  weakly equivalent and $\lambda a. g(f(a)) \equiv \lambda a.a$. They
  are \emph{strongly equivalent} if further the other composition is
  the identity, i.e., $\lambda b. f(g(b)) \equiv \lambda b.b$.
\end{definition}

%% That is, grammars are weakly equivalent if we can construct a parse
%% tree of $B$ from a parse tree of $A$ and vice-versa, whereas they are
%% strongly equivalent if these processes are inverse to each other. The
%% definition of retract is a useful intermediate notion. In retrospect,
%% our axiom about $\StringGram$ says that $\StringGram$ is a retract of $\top$.

@@

%% \paragraph{Ambiguity}
A formal grammar $\A$ is ambiguous if there are multiple parse trees
for the \emph{same} string $\w$.  For example, $\a \oplus \a$ is
ambiguous because there are two parses of $\stringquote{a}$,
constructed using $\inl$ and $\inr$. On the other hand, a formal
grammar is unambiguous when there is at most one parse tree for any
input string. We can capture this notion as a type in \theoryabbv in a
clever way:
\begin{definition}
  \label{def:unambig}
  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
  then $\f \equiv \g$.
\end{definition}
\cref{def:unambig} can be read more intuitively as stating that $\A$
is unambiguous if there is at most one way to transform parses of any
other grammar $\B$ into parses of $\A$. This notion of an unambiguous
type is the analog for linear types of the definition of a (homotopy)
\emph{proposition} in the terminology of homotopy type
theory\cite{hottbook}. The most basic unambiguous types are $\top$ and
$0$, and in a system of classical logic all unambiguous types would
have to be equivalent to one of these, but with our axioms we can show
also that $\I$ and literals $\literal c$ are unambiguous. To see this, first, we
establish two useful properties of unambiguity.
\begin{lemma}[\agdalogo]
  \label{lem:retract}
  If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
\end{lemma}
\begin{lemma}[\agdalogo]
  \label{lem:unambig-sum}
  As a consequence of \cref{cor:binary-mono}, if a binary disjunction $A \oplus B$ is unambiguous then
  $A$ and $B$ are each unambiguous.
\end{lemma}
From \cref{lem:retract}, we can prove that $\StringGram$ is unambiguous,
since it is a retract of $\top$. In fact, observe that if $\A$ is a
retract of $\B$ and $\B$ is unambiguous, then in fact $\A$ and $\B$
are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
\lambda b. b$ follows because $\B$ is unambiguous. Therefore
$\StringGram$ is also strongly equivalent to $\top$. Next, since $\StringGram$ is
defined as a Kleene star, we can easily show that $\StringGram \cong \I
\oplus \CharGram \oplus (\CharGram \otimes \CharGram \otimes \StringGram)$. Then by
\cref{lem:unambig-sum}, we have that $\I$ and
$\CharGram$ --- and thus each literal
$\literal c$ --- are unambiguous as well.

% \paragraph{A Grammar of Strings}
% Define the character grammar $\CharGram$ to be the disjunction over characters in the alphabet $\Sigma$,
% $\CharGram = \LinSigTy {c} {\Sigma} c$. Then define a linear type of strings as a
% linear list of characters,
% $\StringGram = \CharGram^{*}$.

% In the grammars model, we can build a
% strong equivalence between $\top$ and $\StringGram hen building parsers, we may wish to enforce this
% syntactically by including $\top \cong \StringGram$ as an axiom.

% Through inclusion of this axiom, $! : \StringGram \to \top$ is an isomorphism, and
% thus also a monomorphism. That is, $\StringGram$ is unambiguous.
% \max{we should have a section of the previous section which gives the axioms, this one and the distributivity axiom}

% \paragraph{Parseability}
% We say $A$ is \emph{parseable} if there exists a grammar $B$ and a map
% $\top \to A \oplus B$. Taking $\top \cong \StringGram$, a grammar $A$ is parseable
% exactly when we may read an arbitrary string into a parse of $A$ or a $B$-valued
% refutation.

@@

We now turn to our main task, which is using our linear type system to
implement verified parsers. Given a grammar defined as a linear type
$A$, a first attempt at defining a parser would be to implement a
function $\uparrow{(\StringGram \lto A)}$. But since our linear functions
must be total, this means that we can construct an $A$ parse for
\emph{every} input string, which is impossible for most grammars of
interest. Instead we might try to write a partial function as a
$\ltonl{(\StringGram \lto (A \oplus \top))}$ using the ``option''
monad. This allows for the possibility that the input string doesn't
parse, but is far too weak as a specification: we can trivially
implement a parser for any type by always returning $\inr$. The
correct notion of a parser should be one that allows for failure, but
only in the case that a parse cannot be constructed.

\begin{definition}
  \label{def:disjoint}
  Linear types $A$ and $B$ are \emph{disjoint} if there is a function
  $\ltonl {\left( A \& B \lto 0 \right)}$.
\end{definition}

\begin{definition}
  \label{def:parser}
  A parser for a linear type $A$ is the choice a type $A_{\neg}$ disjoint from
  $A$ and function $\uparrow{(\StringGram \lto
    A \oplus A_{\neg})}$.
\end{definition}
Here we replace $\top$ in our partial parser type with a type
$A_{\neg}$ that we can think of as a negation of $A$. The function
$\uparrow{(A \& A_{\neg} \lto 0)}$ ensures that it is impossible for
$A$ and $A_{\neg}$ to parse the same input string. This means that in
defining a parser, we will need to define a kind of negative grammar
for strings that do not parse.

@@

\begin{lemma}[\agdalogo]
  \label{lem:unambig-to-disjoint}
  If $\LinSigTy{x}{X}{A\,x}$ is unambiguous, then for $x \neq x'$, $A\,x$ and $A\,x'$ are
  disjoint. In particular, if the binary product $A \oplus A_{\neg}$ is unambiguous, then
  $A$ and $A_{\neg}$ are disjoint.
\end{lemma}

\begin{proof}
If $\LinSigTy{x}{X}{A\,x}$ is unambiguous, then all functions into it from
$A\,x \& A\,x'$ are equal. In particular,
$\sigma\,x\,\circ\,\pi_{1} \equiv \sigma\,x'\,\circ\,\pi_{2}$ so there is a
function \(
\ltonl{\left(A\,x \& A\,x' \lto
  \equalizer{b}{\left( \sigma\,x \circ \pi_1 \right)}{\left(  \sigma\,x'\circ \pi_2 \right)}
\right)}.
\)
We then compose with the function in \cref{ax:disjointness} to prove that $A\,x$ and $A\,x'$ are disjoint.
\end{proof}

Writing a parser as a linear term is an intrinsic
verification of the \emph{soundness} of the parser completely for free
from the typing: any $\inl$ parse that we return \emph{must}
correspond to a parse tree of the input string. Further, if we verify
the disjointness property of \cref{def:parser} we then also get the \emph{completeness} of
the parser as well, that it never fails to generate an $A$ parse when
it is possible.

Our main method for constructing verified parsers is to show that a
grammar $A$ is weakly equivalent to a grammar for a deterministic
automaton. Parsers for deterministic automata are simple to implement
by stepping through the states of the automaton, with the rejecting
traces serving as the negative grammar. This is sufficient due to the
following:
\begin{lemma}[\agdalogo]
  \label{lem:wk-eqv-parse}
  If $\A$ is weakly equivalent to $\B$ then any parser for $\A$ extends to a
  parser for $\B$.
\end{lemma}
Here we need both directions of the weak equivalence. We need $A \lto
B$ to extend the parser from $\StringGram \lto A \oplus A_{\neg}$ to
$\StringGram \lto B \oplus A_{\neg}$ but then we also need $B \lto A$ to
establish that $A_{\neg}$ is disjoint from $B$.

@@

@@

Because DFAs are deterministic, we are able to prove that their type
of traces are unambiguous and define a parser for them. In particular
we show that for any start state $s$, ${\LinSigTy {b} {\Bool}
  \Trace_{D}~s~b}$ is a retract of $\StringGram$ and apply \cref{lem:retract} to
derive unambiguity. That is, first we
construct a function $\pparse_{D}$ that is a parser for
$\Trace_D~{s}~{true}$, with $\Trace_D~{s}~{false}$ being the disjoint
type used. The fact that these trace types are disjoint follows from the
unambiguity of $\LinSigTy {b} {\Bool} {\Trace_{D}~s~b}$ by \cref{lem:unambig-sum}.

The parser,
$\pparse_{D}$,
is defined by recursion on strings in \cref{fig:printdfa}. If this string is empty, then $\pparse_{D}$
defines a linear function that terminates a trace at the input state $s$. If the
string is nonempty, then $\pparse_{D}$ walks forward in $D$ from the input state $s$ by
the character at the head of the string.
%
The inverse, $\print_{D}$ is defined by recursion on traces. If the trace is defined
via $\nil$, then $\print_{D}$ returns the empty string. Otherwise, if the trace
is defined by $\cons$ then $\pparse_{D}$ appends the character from the most
recent transition to the output
string and recurses.
%
We prove this is a retraction by induction on traces.

\begin{theorem}[\agdalogo]
  \label{thm:dfa-parser}
  $\pparse_D s$ is a parser for $\Trace_D~s~\true$.
\end{theorem}

@@

To construct a term from DFA traces to NFA traces, we similarly strengthen our
inductive hypothesis and build a parse transformer
\[
  \DtoN : \ltonl {\left( \Trace_{\D}~\X~\true \lto \bigoplus\limits_{s : \N.\states}\bigoplus\limits_{\texttt{sInX} : \X \ni \s} \Trace_{\N}~s~\true \right)}
\]

Because the states of $D$ are $\epsilon$-closed subsets of $N.\states$, if two
states $s$ and $s'$ belong to the the same $\epsilon$-closed subset
$X : \mathbb{P}_{\epsilon}(N.\states)$ then there exists a $\epsilon$-path in
$N$ between the two, but we do not necessarily know which one. Similarly, the
data contained in the type $\Trace_{\D}}~X~true$ is the \emph{existence} of an accepting trace in
$\N$ beginning at some state in $\D$.

To define $\DtoN$, we need a choice function that extracts out a trace in $N$
from the mere existence of one. We achieve this by choosing the smallest trace
through $\N$ subject to an ordering on the non-linear types $N.\states$,
$N.\transitions$, and $N.\epstransitions$. In essence, the given orderings
specify a global disambiguation strategy.

@@

\subsection{Unrestricted Grammars}
